---
title: "Dataframes"
author: "Dr. Shoemaker"
date: "DATA 2401"
output: slidy_presentation
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#library(tidyverse) # loading packages at the beginning is a good idea
```

## What are data frames? 

A data frame is essentially a table for your data! 

They're really just a conglomeration of vectors into one object, where every vector, aka every column, has to have the same length. 

So now you can have a table where one column is a column of names, then a column of dates, then a column of salaries. 

## Making Data frames

You'll often load in data frames from an external source, but it's also useful to know how to make them yourself. 

The function `data.frame()` takes vectors as arguments, and creates a table with a column for each vector. 

```{r}

# a vector of names
name <- c("Ada", "Bob", "Chris", "Daisy", "Elda")

# a vector of salaries
salary <- c(95, 82, 75, 92, 102)

# a vector of scores
score <- c(6, 4, 2, 5, 9)

# Combine into a dataframe
# The names of the variables become the names of the columns! 
people <- data.frame(name, salary, score, stringsAsFactors = F)
people
```


# A Note on Factors



- In a recent(ish) R update, it went from default `T` to default `F`, so you really don't have to worry about factors when you read in data, just when you want to use them.
   + you can see what the default for your machine is by running `default.stringsAsFactors()`
- This tells R that you want the strings to be kept as strings, and not as "factors", a different data type that's often used for building statistical models. 
  + If your strings are for a categorical variable, you *want* to have them as factors. 

  
Just keep this in mind! You will occasionally need to swap back and forth, using `as.factor()` and `as.character()` 

```{r}
a <- c("a", "a", "b", "b")
str(a)
a

a2 <- as.factor(a)
str(a2)
a2

a3 <- as.character(a2)
str(a3)
a3
```




# Describing Structure

Dataframes have properties you may want to use or reference. 

For starters, the dimensions, the number of rows and columns.  

- For example, the number of columns = the number of variables, which is a parameter in some statistical models

```{r}
# to get rows and columns:
nrow(people)
ncol(people)

# to get both
dim(people) # Note this returns a vector! 
```


---

- The names of the columns and the rows:

```{r}
colnames(people)
rownames(people)
```

If you want to change the names of the columns or rows, you just need to assign the new names to that function! 

```{r}
rownames(people) <- people$name
colnames(people) <- c("Name", "Salary", "Score")
people
```

Note: It's typical practice to not use rownames, but rather have a variable that contains the identifying feature.

To get rownames to a column, we have a function, `rownames_to_columns()`, or you can just save it as an new element to the dataframe! 

Note: This is our first example of adding a column to a dataframe, and our first time using the dollar sign indexing. 

```{r}
# need to load tidyverse to get access to the function.

people <- rownames_to_column(people, var = "ID")
people
# OR 
people$id <- rownames(people)

```

--- 

### Peeking at your data. 

Different ways of looking at "glimpses" of the data frame: 

One of the quickest is `glimpse`, but we can also use several different options.

```{r}
library(palmerpenguins)

  # load in a library for access to fun penguin data
library(dplyr) # or library(tidyverse), which we've already loaded !
  # load in a library for access to our more "data science" functions

glimpse(penguins)

# For these, you can set the number of rows you want to see, or leave it blank
head(penguins) # shows the first 6 rows, by default
head(penguins, 2) # shows the first 2 rows
tail(penguins, 2) # shows the last 2 rows

 
slice(penguins, c(1,4)) # shows the rows you pass as the second argument


# View(penguins) # in RStudio, pulls up a new "tab" with the data
  ### Doesn't work when you're knitting on a Mac, so I've commented it out...
```



# Referencing Data

We can use these dollar sign and double bracket notations to get whole vectors from the data frame.

```{r}
people$Name

people[['Score']]

penguins$island

```


OR we can use single bracket notations to access more flexibly. 

`data_frame_name[row_index, column_index]`    


```{r}
# the third element of the second column
people[3, 2] 

# all elements of the second column
people[ , 2]

# the third row
people[3, ]

# the column called 'Score'
people[ , 'Score']

# the second element in the Score variable
people[2, 'Score']
people['Chris', 'Score']
```

---

We can combine this with vectors and logical operators to get just what we want! 

```{r}
people[2:4, ] # rows 2, 3 and 4

people[ ,c('Name','Salary')] # just those two columns

# all the columns for the rows where the salary value is greater than 90
people[people$Salary > 90, ] 

# the penguins on Torgersen island
penguins[penguins$island == "Torgersen", ]
```

In general, it is safer to reference by variable name. 

```{r, eval = F}

penguins$species
### is better than 
penguins[ , 1]
```

Consider if you add an additional column, or they get changed around in order. "Hard coding" the specific row or column you want can lead to errors. 



# Exercise: 

- Create a vector of the number of points the Dash scored in the first 4 games
of the last season (google "Dash" for the scores!).

```{r}
points = c(1,0,3,1)
points_allowed = c(5,0,2,3)

games = data.frame(points,points_allowed)
games

```

- Create a vector of the number of points the Dash have allowed to be scored
against them in each of the first 4 games of the season.

```{r}

```


- Combine your two vectors into a dataframe called `games`.

```{r}


```

- Create a new column "diff" that is the difference in points between the teams
Hint: recall the syntax for assigning new elements (which in this case will be
a vector) to a *list*! It's similar.

```{r}
points = c(1,0,3,1)
points_allowed = c(5,0,2,3)

games = data.frame(points,points_allowed)
games
games$diff <- c(-4,0,1,-2)
games

```

- Create a new column "won" which is TRUE if the Dash won the game. (Use logical operators).

```{r}
points = c(1,0,3,1)
points_allowed = c(5,0,2,3)

games = data.frame(points,points_allowed)
games
games$diff <- c(-4,0,1,-2)
games$won <- games$diff >0
games

```

- Create a vector of the opponent's names corresponding to the games played.

```{r}
points = c(1,0,3,1)
points_allowed = c(5,0,2,3)

games = data.frame(points,points_allowed)

games$diff <- c(-4,0,1,-2)
games$won <- games$diff >0
games$opponent <- c("North Carolina Courage","Racing Louisville","Bay_FC","Washington Spirit" )

```

- Assign that vector to a new column of the dataframe. 

```{r}
points = c(1,0,3,1)
points_allowed = c(5,0,2,3)

games = data.frame(points,points_allowed)

games$diff <- c(-4,0,1,-2)
games$won <- games$diff >0
games$opponent <- c("North Carolina Courage","Racing Louisville","Bay_FC","Washington Spirit" )
games

```

- View your data frame to see how it looks!

```{r}
points = c(1,0,3,1)
points_allowed = c(5,0,2,3)

games = data.frame(points,points_allowed)

games$diff <- c(-4,0,1,-2)
games$won <- games$diff >0
games$opponent <- c("North Carolina Courage","Racing Louisville","Bay_FC","Washington Spirit" )
games
```


# Working with Built-in R data 

There are lots of data frames built into "base" R, as well as available in packages. 

Go down to your console, type `data()` and hit tab. All of these are data that available to you right now in R. 

Some are matrices, that you will need to turn into dataframes! You do this with `as.data.frame()` 

You'll work more with these in today's lab. 

# Working with CSV data! 

One of the most common file formats you're going to come across is the Comma Separated Value format, `.csv`.

Programs like Excel and Google Sheets can work with csv files, but it's important to remember that any sort of formulas or formatting you do on the csv file in those programs do NOT get saved. Only the data. 

---

If you open a csv in a program like Atom or Sublime, it will look like this: 

```
Col_1,Col_2,Col_3
Abc,1,2.3
Def,2,4.5
Ghi,3,6.7
```

Literally, **comma separated values**. 

- In general, different types of files use different types of separators. 
- You won't see these separators if you open the file in excel, as excel translates "," into "the next cell" 
- You can always open them in an editor like Sublime to see what's going on.
- If the file has a different type of separator than a comma, you can specify this in `read.csv()` with the argument `sep = `


---

To read in a csv file, you use the function `read_csv()` 

> This function is in the package `readr`, which is in the tidyverse. If your code below doesn't work, make sure you ran the `library(tidyverse)` at the top of the file, or run `library(readr)` now.

```{r}
loaded_data <- read_csv('../Data/read_in_test.csv')

# if you save it in the same folder
# loaded_data <- read_csv("read_in_test.csv")

# if you don't know where you saved this file or that one, use the full path: 
# loaded_data <- read_csv("~/OneDrive - University of Houston Downtown/0_Teaching/Courses/DATA2401/Data/read_in_test.csv")
# this works for my machine. It will not work for yours.
```

Note that you can either use a full path here, or the relative path from your current working directory. 

# Other file types to read in?

- `read_csv()` reads comma delimited files 
- `read_csv2()` reads semicolon separated files (common in countries where , is used as the decimal place)
- `read_tsv()` reads tab delimited files 
- `read_delim()` reads in files with any delimiter.

> There are also extra packages for specific things. For example, `readxl` has specific functions to read in Excel files which have arguments where you can choose which Sheet to read in. 

---

You can do the reverse as well! Let's add a column to the data frame and save it back as another csv file. 

```{r}
loaded_data$Col4 <- c(4,5,6)


write_csv(loaded_data, "write_out_test2.csv")
```


# Note: Working directory

When you knit the .Rmd file, your working directory is the place the .Rmd (and html) is saved.

When you're working in the editor in RStudio while **not** in a project (whether on .Rmd or .R files), the working directory isn't automatically that location. 

When you work inside an RProject, the working directory is automatically set to the location of the files. 

- You can find out your current working directory with `getwd()` or More -> Go to working directory in the Files pane
- You can set your working directory with `setwd()` or More -> Set as working directory in the Files pane, when you're "at" the location you want to set, or just by right clicking on the name of the file and setting it there.


# Example: Gates Foundation Educational Grants

Head to the in class repository for an example. 

